A comparison of design-based and model-based approaches for finite population spatial data.

### 2022-04-14 inst/manuscript/manuscript.Rmd
section 4 updates

We define the relative RMSE as a ratio with numerator given by the RMSE for a sampling-inference combination and the denominator given by the RMSE for SRS-DB. Relative RMSEs for the random location layout are provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 0%"), the four sampling-inference combinations have approximately equal RMSE. In these scenarios, using GRTS sampling or model-based inference does not generally increase efficiency compared to SRS-DB. When there is spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 0%" and "DRE%: 50%"), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB, SRS-MB, and finally SRS-DB, though the difference in relative RMSE among GRTS-MB, GRTS-DB, and SRS-MB is small.  As the strength of spatial covariance increases, the gap in RMSE between SRS-DB and the other sampling-inference combinations widens. Finally we note that when there is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB, suggesting that the lack of efficiency from SRS is largely mitigated by model-based inference. These RMSE conclusions are similar to those observed in the grid location layout, so we omit a figure and discussion regarding the grid location layout here. Tables for RMSE in all 36 simulation scenarios are provided in the supporting information.
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes}
  \label{fig:rmspe_eff}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/data_rmspe_eff.jpeg}
  \caption{Relative RMSE in the data study for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:data_rmspe_eff}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/data_coverage.jpeg}
  \caption{Interval coverage in the data study for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black line represents 95\% coverage.}
  \label{fig:data_figconf}
\end{figure}

@@

The design-based and model-based approaches to statistical inference are fundamentally different paradigms. The design-based approach relies on random sampling to estimate population parameters. The model-based approach relies on distributional assumptions to predict realized values of a stochastic process. Though the model-based approach does not rely on random sampling, it can still be beneficial as a way to guard against preferential sampling. While the design-based and model-based approaches have often been compared in the literature from theoretical and analytical perspectives, our contribution lies in studying them in a spatial context while implementing spatially balanced sampling and the design-based, local neighborhood variance estimator. Aside from the theoretical differences described, a few analytical findings from the simulation study are particularly notable. First, independent of the analysis approach, we found no reason to prefer IRS over GRTS when sampling spatial data -- GRTS-DB and GRTS-MB generally had similar RMSE as their IRS counterparts when there was no spatial covariance and lower RMSE than their IRS counterparts when there was spatial covariance. Second, the sampling decision (IRS vs GRTS) is most important when using a design-based analysis. Though GRTS-MB still had lower RMSE than SRS-MB, the model-based analysis mitigated most of the RMSE inefficiencies that result from the IRS samples lacking spatial balance. Third, as the strength of spatial covariance increases, the gap in RMSE and MStdE between SRS-DB and the other sampling-inference combinations also increases, likely because SRS-DB is the only combination that ignores spatial locations in sampling and analysis. Fourth and finally, when the response was normal, interval coverage for all sampling-inference combinations was usually close to 95% for all sample sizes; when the response was lognormal, interval coverage for all sampling-inference combinations was usually between 90% and 95% and closest to 95% when $n = 200$.

AT LEAST HAVE DISCUSSION ABOUT MODEL BASED ASSUMPTIONS AND MOVE VALIDITY COMMENTS TO RESULTS SECTION.

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discussed, but others we have not, and they are worthy of consideration in future research.  Design-based approaches are often computationally efficient, while model-based approaches can be computationally burdensome, especially for likelihood-based estimation methods like REML that rely on inverting a covariance matrix. The design-based approach easily handles binary data through a straightforward application of the Horvitz-Thompson estimator. In contrast, analyzing binary data using a model-based approach generally requires a logistic mixed regression model, which can be challenging to estimate and interpret [@bolker2009generalized]. The design-based approach yields valid results because the sampling plan and inclusion probabilities are specified directly by the researcher, while the model-based approach may not yield valid results if the assumptions made do not not accurately capture reality. The model-based approach, however, can more naturally quantify the relationship between covariates (predictor variables) and the response variable. The model-based approach also yields estimated spatial covariance parameters, which help better understand the dependence structure in the process in study. Model selection is also possible using model-based approaches and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based approaches are capable of more efficient small-area estimation than design-based approaches by leveraging distributional assumptions in areas with few observed units. Model-based approaches can also compute unit-by-unit predictions at unobserved locations and use them to construct informative visualizations like smoothed maps. @brus1997random provide a more thorough discussion regarding the benefits and drawbacks of the two approaches. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.

>>>>

We define the relative RMSE as a ratio with numerator given by the RMSE for a sampling-inference combination and the denominator given by the RMSE for SRS-DB. Relative RMSEs for the random location layout are provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 0%"), the four sampling-inference combinations have approximately equal RMSE. In these scenarios, using GRTS sampling or model-based inference does not generally increase efficiency compared to SRS-DB. When there is spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 0%" and "DRE%: 50%"), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB, SRS-MB, and finally SRS-DB, though the difference in relative RMSE among GRTS-MB, GRTS-DB, and SRS-MB is small.  As the strength of spatial covariance increases, the gap in RMSE between SRS-DB and the other sampling-inference combinations widens. Finally we note that when there is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB, suggesting that the lack of efficiency from SRS is largely mitigated by model-based inference. These RMSE conclusions are similar to those observed in the grid location layout, so we omit a figure and discussion regarding the grid location layout here. Tables for RMSE in all 36 simulation scenarios are provided in the supporting information.
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:rmspe_eff}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/data_rmspe_eff.jpeg}
  \caption{Relative RMSE in the data study for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:data_rmspe_eff}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/data_coverage.jpeg}
  \caption{Interval coverage in the data study for the four sampling-inference combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.The solid black lines separate the sample sizes and the dashed black lines represent 95\% coverage.}
  \label{fig:data_figconf}
\end{figure}

@@

The design-based and model-based approaches to statistical inference are fundamentally different paradigms. Design-based inference relies on random sampling to estimate population parameters. Model-based inference relies on distributional assumptions to predict realized values of a data-generating stochastic process. Though model-based inference does not rely on random sampling, it can still be beneficial as a way to guard against preferential sampling. While design-based inference and model-based inference have often been compared in the literature from theoretical and analytical perspectives, our contribution lies in studying them for finite population spatial data while implementing GRTS sampling and the local neighborhood variance estimator. Aside from the theoretical differences described throughout the manuscript, a few analytical findings from the simulated and real data studies were particularly notable. All sampling-inference combinations had approximately zero mean bias. Independent of the inference approach, GRTS-DB and GRTS-MB had lower RMSE than their SRS counterparts. Though GRTS-DB and GRTS-MB generally had very similar RMSE, SRS-MB tended to have much lower RMSE than SRS-DB, suggesting that the model-based inference mitigated much of the inefficiency in RMSE from SRS. As the proportion of dependent random error in the simulated data increased, SRS-MB, GRTS-DB, and GRTS-MB become increasingly more efficient (lower RMSE) than SRS-DB. Interval coverage tended to be higher for the symmetric responses than skewed responses and tended to increase with the sample size. At a sample size of $n = 200$, generally all interval coverages were near the desired value of 95%.

There are several benefits and drawbacks of the design-based and model-based approaches for finite population spatial data. Some we have discussed, but others we have not, and they are worthy of consideration in future research.  First we discuss advantages of design-based inference. Design-based inference is often computationally efficient, while model-based inference can be computationally burdensome, especially for likelihood-based estimation methods like REML that rely on inverting a covariance matrix. Design-based inference easily handles binary data through a straightforward application of the Horvitz-Thompson estimator. In contrast, analyzing binary data using model-based inference generally requires a logistic mixed regression model, which can be difficult to estimate and interpret [@bolker2009generalized]. An advantage of design-based inference is that interval coverage is valid (has the proper coverage rate) as long as 1) the sample is sufficiently large to ensure the statistic's sampling distribution is approximately normal and 2) the variance estimator is consistent [@brus1997random; @sarndal2003model]. This is because with design-based inference, the sampling plan and inclusion probabilities are specified directly by the researcher. With the model-based approach, however, interval coverage is unlikely to be valid if the model's assumptions made do not not accurately reflect reality. Whether a model's assumptions accurately reflect reality can be a challenging and sometimes impossible question to answer definitively. Now we discuss advantages of model-based inference. Model-based inference can more naturally quantify the relationship between covariates (predictor variables) and the response variable than design-based inference. Model-based inference also yields estimated spatial covariance parameters, which help better understand the dependence structure of the process in study. Model selection is also possible using model-based inference and criteria such as cross validation, likelihood ratio tests, or AIC [@akaike1974new]. Model-based inference is capable of more efficient small-area estimation than design-based inference because model-based inference can leverage distributional assumptions in areas with few observed population units. Model-based inference can also compute unit-by-unit predictions at unobserved locations and use them to construct informative visualizations like smoothed maps. @brus1997random provide a more thorough discussion regarding the benefits and drawbacks of the two approaches. In short, when deciding whether the design-based or model-based approach is more appropriate to implement, the benefits and drawbacks of each approach should be considered alongside the particular goals of the study.

There are many extensions of this research worthy of future consideration, some of which we discuss next. Some extensions to study include sampling with unequal inclusion probabilities, different spatially balanced sampling approaches (instead of GRTS), different spatial data configurations, different spatial domains like stream networks [@ver2010moving], different response or covariance structures, the effect of spatial or external mean trends (which can be defined through covariates), and more.

### 2022-04-14 inst/manuscript/manuscript.tex
section 4 updates

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes}
  \label{fig:rmspe_eff}
\end{figure}

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach relies on
random sampling to estimate population parameters. The model-based
approach relies on distributional assumptions to predict realized values
of a stochastic process. Though the model-based approach does not rely
on random sampling, it can still be beneficial as a way to guard against
preferential sampling. While the design-based and model-based approaches
have often been compared in the literature from theoretical and
analytical perspectives, our contribution lies in studying them in a
spatial context while implementing spatially balanced sampling and the
design-based, local neighborhood variance estimator. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, independent of the
analysis approach, we found no reason to prefer IRS over GRTS when
sampling spatial data -- GRTS-DB and GRTS-MB generally had similar RMSE
as their IRS counterparts when there was no spatial covariance and lower
RMSE than their IRS counterparts when there was spatial covariance.
Second, the sampling decision (IRS vs GRTS) is most important when using
a design-based analysis. Though GRTS-MB still had lower RMSE than
SRS-MB, the model-based analysis mitigated most of the RMSE
inefficiencies that result from the IRS samples lacking spatial balance.
Third, as the strength of spatial covariance increases, the gap in RMSE
and MStdE between SRS-DB and the other sampling-inference combinations
also increases, likely because SRS-DB is the only combination that
ignores spatial locations in sampling and analysis. Fourth and finally,
when the response was normal, interval coverage for all
sampling-inference combinations was usually close to 95\% for all sample
sizes; when the response was lognormal, interval coverage for all
sampling-inference combinations was usually between 90\% and 95\% and
closest to 95\% when \(n = 200\).

AT LEAST HAVE DISCUSSION ABOUT MODEL BASED ASSUMPTIONS AND MOVE VALIDITY
COMMENTS TO RESULTS SECTION.

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. Design-based approaches are often computationally
efficient, while model-based approaches can be computationally
burdensome, especially for likelihood-based estimation methods like REML
that rely on inverting a covariance matrix. The design-based approach
easily handles binary data through a straightforward application of the
Horvitz-Thompson estimator. In contrast, analyzing binary data using a
model-based approach generally requires a logistic mixed regression
model, which can be challenging to estimate and interpret (Bolker et
al., 2009). The design-based approach yields valid results because the
sampling plan and inclusion probabilities are specified directly by the
researcher, while the model-based approach may not yield valid results
if the assumptions made do not not accurately capture reality. The
model-based approach, however, can more naturally quantify the
relationship between covariates (predictor variables) and the response
variable. The model-based approach also yields estimated spatial
covariance parameters, which help better understand the dependence
structure in the process in study. Model selection is also possible
using model-based approaches and criteria such as cross validation,
likelihood ratio tests, or AIC (Akaike, 1974). Model-based approaches
are capable of more efficient small-area estimation than design-based
approaches by leveraging distributional assumptions in areas with few
observed units. Model-based approaches can also compute unit-by-unit
predictions at unobserved locations and use them to construct
informative visualizations like smoothed maps. Brus and De Gruijter
(1997) provide a more thorough discussion regarding the benefits and
drawbacks of the two approaches. In short, when deciding whether the
design-based or model-based approach is more appropriate to implement,
the benefits and drawbacks of each approach should be considered
alongside the particular goals of the study.

@@

>>>>

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:rmspe_eff}
\end{figure}

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. Design-based inference relies on
random sampling to estimate population parameters. Model-based inference
relies on distributional assumptions to predict realized values of a
data-generating stochastic process. Though model-based inference does
not rely on random sampling, it can still be beneficial as a way to
guard against preferential sampling. While design-based inference and
model-based inference have often been compared in the literature from
theoretical and analytical perspectives, our contribution lies in
studying them for finite population spatial data while implementing GRTS
sampling and the local neighborhood variance estimator. Aside from the
theoretical differences described throughout the manuscript, a few
analytical findings from the simulated and real data studies were
particularly notable. All sampling-inference combinations had
approximately zero mean bias. Independent of the inference approach,
GRTS-DB and GRTS-MB had lower RMSE than their SRS counterparts. Though
GRTS-DB and GRTS-MB generally had very similar RMSE, SRS-MB tended to
have much lower RMSE than SRS-DB, suggesting that the model-based
inference mitigated much of the inefficiency in RMSE from SRS. As the
proportion of dependent random error in the simulated data increased,
SRS-MB, GRTS-DB, and GRTS-MB become increasingly more efficient (lower
RMSE) than SRS-DB. Interval coverage tended to be higher for the
symmetric responses than skewed responses and tended to increase with
the sample size. At a sample size of \(n = 200\), generally all interval
coverages were near the desired value of 95\%.

There are several benefits and drawbacks of the design-based and
model-based approaches for finite population spatial data. Some we have
discussed, but others we have not, and they are worthy of consideration
in future research. First we discuss advantages of design-based
inference. Design-based inference is often computationally efficient,
while model-based inference can be computationally burdensome,
especially for likelihood-based estimation methods like REML that rely
on inverting a covariance matrix. Design-based inference easily handles
binary data through a straightforward application of the
Horvitz-Thompson estimator. In contrast, analyzing binary data using
model-based inference generally requires a logistic mixed regression
model, which can be difficult to estimate and interpret (Bolker et al.,
2009). An advantage of design-based inference is that interval coverage
is valid (has the proper coverage rate) as long as 1) the sample is
sufficiently large to ensure the statistic's sampling distribution is
approximately normal and 2) the variance estimator is consistent (Brus
and De Gruijter, 1997; Särndal et al., 2003). This is because with
design-based inference, the sampling plan and inclusion probabilities
are specified directly by the researcher. With the model-based approach,
however, interval coverage is unlikely to be valid if the model's
assumptions made do not not accurately reflect reality. Whether a
model's assumptions accurately reflect reality can be a challenging and
sometimes impossible question to answer definitively. Now we discuss
advantages of model-based inference. Model-based inference can more
naturally quantify the relationship between covariates (predictor
variables) and the response variable than design-based inference.
Model-based inference also yields estimated spatial covariance
parameters, which help better understand the dependence structure of the
process in study. Model selection is also possible using model-based
inference and criteria such as cross validation, likelihood ratio tests,
or AIC (Akaike, 1974). Model-based inference is capable of more
efficient small-area estimation than design-based inference because
model-based inference can leverage distributional assumptions in areas
with few observed population units. Model-based inference can also
compute unit-by-unit predictions at unobserved locations and use them to
construct informative visualizations like smoothed maps. Brus and De
Gruijter (1997) provide a more thorough discussion regarding the
benefits and drawbacks of the two approaches. In short, when deciding
whether the design-based or model-based approach is more appropriate to
implement, the benefits and drawbacks of each approach should be
considered alongside the particular goals of the study.

There are many extensions of this research worthy of future
consideration, some of which we discuss next. Some extensions to study
include sampling with unequal inclusion probabilities, different
spatially balanced sampling approaches (instead of GRTS), different
spatial data configurations, different spatial domains like stream
networks (Ver Hoef and Peterson, 2010), different response or covariance
structures, the effect of spatial or external mean trends (which can be
defined through covariates), and more.

@@

\leavevmode\hypertarget{ref-ver2010moving}{}%
Ver Hoef, J.M., Peterson, E.E., 2010. A moving average approach for
spatial statistical models of stream networks. Journal of the American
Statistical Association 105, 6--18.
